Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-09-11 manuscript/manuscript.Rmd
1st trial of 2nd round revision.

---
title: "Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
abstract: |
  ```{=latex}
  This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology. The Triple P framework emphasizes the amplification of multi-level variation and the identification of interacting relationships of variations across the levels of perception, process, and product. Here I propose the following three basic measures to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.

@@

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology. Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here, I believe, provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* for most social scientists in the post-Kahneman era, but particularly anthropologists and archaeologists [@henrichSearchHomoEconomicus2001].

As such, built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations. To accomplish these two objectives, I advocate the following three measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potential of this framework.

# What good is actualistic experiment?

The trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267], and the past decades have witnessed experimental archaeology's growing research interests focusing on the causal mechanism at the behavioral level in the explanation of material culture variation [@eren2016; @lin2018]. In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.). All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design is incapable of inferring how context-generic principles interact in a particular context as reflected in real-world conditions. In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies, which was demonstrated to be problematic [@liu2023]. In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of skill acquisition. After all, these experimental results can only be as robust as their experimental settings.

More importantly, controlled experiments without randomization are not enough to infer causal mechanisms, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012]. This issue is rarely discussed in material science oriented experiments but should raise a critical concern in experimental archaeology as social science. Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved. Rather, most of our knowledge regarding the past is derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through experiments featuring small-sized convenience sample. It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the gold standard of knowledge in the realm of philosophy of science [@cartwright2007].

@@

Modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century, rather than from the inter-generational transmission of knapping knowledge spanning millions of years. This historical context is well elucidated by studies on the subject [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61]. Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers. They tend to come up with ideas that may appear to be counter-intuitive at first glance for academics. One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree[-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology. In his experiment, Crabtree demonstrated the excellent performance of blade dorsal ridge on tasks like shaving and cutting hard materials, challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking and shedding light on future functional reconstruction through the use-wear analysis. It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] famous ethnography. Likewise, novices' lack of expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers. Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts.

@@

As implied in its name, the implementation of the Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265]. Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends. While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentions simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage [@reynolds1999; @stout2023]. The importance of specifying and documenting the context information of both the experiment as well as the phenomenon of interest has also been recently highlighted in psychological sciences [@holleman2020].

@@

Traditionally speaking, the product-level data, namely the documentation and analysis of replicas, form the sole research subject of experimental archaeology and serve as the tangible foundation for analogical inference in the interpretation of archaeological materials. It can exist in the form of spreadsheets containing detailed technological attributes, or photos and illustrations, or high-resolution 3D scans of individual artifact or a whole assemblage. No particular modification regarding the collection procedure of product-level data is required in the context of the Triple P framework, although it should be emphasized that the definition of variables measured and the documentation techniques (models of camera/scanners, light setting, processing software version and workflow, etc.) should be always available in the relevant meta-data. Besides, adopting good habits in spreadsheet data organization is strongly recommended [@broman2018].

@@

While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent. One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023]. Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal. This approach is a powerful and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handle variation very well. To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner. However, this may be biased given that 1) experiments and modeling studies in cognitive science have shown that novices often feature a low planning depth [@vanopheusden2023] and 2) ethnographic studies demonstrated that even expert practitioners in traditional societies can have a different set of perception on the causal structure of how certain behaviors will modify the raw materials [@harris2021].

Consequently, we need to accumulate more real-world data by recording a large amount videos of toolmaking and conducting systematical ethogram analysis. With the emergence of new software platforms such as BORIS [@friard2016], the difficulty of coding has decreased significantly in recent years (**Figure** \@ref(fig:ethogram)). Here I use a modified version of action grammar developed by [@stout2021] as an example, among multiple coding schemes featuring different research focus [@muller2023]or degrees of coding complexity [@mahaney2014; @cueva-temprana2019]. The knapping action recorded in videos can be coded following the ethogram presented in **Table** \@ref(tab:tab1). Depending on the original research question, sequences of coded actions can then be used in further analysis, such as complexity [@stout2021], similarity [@mobbs2021], etc.

@@

Ethnographies revolving around general archaeological practices [@ethnogra2006], experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including both WEIRD [@whittaker2004] and non-WEIRD populations[@stout2002; @arthur2018], are far from novel. However, it has never been formally recognized as a legitimate research method in experimental archaeology. Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology. This also echoes the post-positivist turn in psychology, a field that is known for the development of experimental methods, in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022; @weger2019; @stoutCognitiveScienceTechnology2021].

Through participant observation, interviews, and detailed field notes, ethnographers can capture the subtle nuances of perception, such as sensory experiences, social interactions, and cultural meanings associated with the experimental activities. Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction. One potential application of ethnographic methods in experimental archaeology of stone artifacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.

@@

The comparative study and large-scale synthesis of variation data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology and likely won't be available in the near future. Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the highest file size compared with the other two. As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library that was originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.

>>>>

---
title: "Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; cheng.liu@emory.edu]
abstract: |
  ```{=latex}
  This paper presents the outline of the Perception-Process-Product ("Triple P") conceptual framework that aims to expand the scope of experimental archaeology. The Triple P framework emphasizes the amplification of multi-level variation and the identification of interacting relationships of variations across the levels of perception, process, and product. Here I propose the following three ontologically and epistemologically based principles to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.

@@

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology. Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* for most anthropologists [@henrichSearchHomoEconomicus2001].

Built upon the *Homo economicus* critique and early works in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations. To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework. Ontologically speaking, it requires 1) acknowledging the contribution and limitations of actualistic experiments and 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices. These two principles are developed to advocate a pluralistic approach to the explanation of complex variation instead of treating the optimization-based research agenda as a panacea. Epistemologically speaking, the Triple P framework 3) adopts a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potential of this framework.

# What good is actualistic experimentation?

The trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267]. The past decades have witnessed experimental archaeology's growing research interests focusing on the causal mechanism at the behavioral level in the explanation of material culture variation [@eren2016; @lin2018]. In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.). All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design is incapable of inferring how context-generic principles interact in a particular context as reflected in real-world conditions. In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies, which was demonstrated to be potentially problematic [@liu2023]. In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of cultural transmission. After all, these experimental results can only be as robust as their experimental settings.

@@

Modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61], rather than from the inter-generational transmission of knapping knowledge spanning millions of years. Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers. They tend to come up with ideas that may appear to be counter-intuitive at first glance for academics. One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree[-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology. In his experiment, Crabtree demonstrated the excellent performance of blade dorsal ridge on tasks like shaving and cutting hard materials, challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking and shedding light on future functional reconstruction through the use-wear analysis. It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] famous ethnography. Likewise, novices' lack of expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers. Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts.

@@

As implied in its name, the implementation of the Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference. Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends. While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentions simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage [@reynolds1999; @stout2023]. The importance of specifying and documenting the context information of both the experiment as well as the phenomenon of interest has also been recently highlighted in psychological sciences [@holleman2020].

@@

Traditionally speaking, the product-level data, namely the documentation and analysis of replicas, form the sole research subject of experimental archaeology and serve as the tangible foundation for analogical inference in the interpretation of archaeological materials. It can exist in the form of spreadsheets containing detailed technological attributes, or photos and illustrations, or high-resolution 3D scans of individual artifact or a whole assemblage. No particular modification regarding the collection procedure of product-level data is required in the context of the Triple P framework, although it should be emphasized that the definition of variables measured and the documentation techniques (models of camera/scanners, light setting, processing software version and workflow, etc.) should be always available in the relevant meta-data. Adopting good habits in spreadsheet data organization is also strongly recommended [@broman2018].

@@

While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely neglected among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent. One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023]. Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal. This approach is a powerful and elegant yet limited by its normative and analytical orientation, meaning it cannot handle variation very well. To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner. However, this may be biased given that 1) experiments and modeling studies in cognitive science have shown that novices often feature a low planning depth [@vanopheusden2023] and 2) ethnographic studies demonstrated that even expert practitioners in traditional societies can have a different set of perception on the causal structure of how certain behaviors will modify the raw materials [@harris2021].

Consequently, we need to accumulate more real-world data by recording a large amount videos of toolmaking and conducting systematical ethogram analysis. With the emergence of new software platforms such as BORIS [@friard2016], the difficulty of coding has decreased significantly in recent years (**Figure** \@ref(fig:ethogram)). Here I use a modified version of action grammar developed by [@stout2021] as an example, among multiple coding schemes featuring different research focus [@muller2023] or granularity [@mahaney2014; @cueva-temprana2019; @roux2005]. The knapping action recorded in videos can be coded following the ethogram presented in **Table** \@ref(tab:tab1). Depending on the original research question, sequences of coded actions can then be used in further analysis, such as complexity [@stout2021], similarity [@mobbs2021], etc.

@@

Ethnographies revolving around experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including contemporary U.S. hobbyists [@whittaker2004] and knapping practitioners in various non-industrial societies [@stout2002; @arthur2018], are far from novel. However, ethnography has never been formally recognized as a legitimate research method in experimental archaeology. Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information (e.g., intention) that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology. This also echoes the post-positivist turn in psychology, a field that is known for the development of experimental methods, in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022; @weger2019; @stoutCognitiveScienceTechnology2021].

Through participant observation, interviews, and detailed field notes, ethnography can capture the subtle nuances of perception, such as sensory experiences, social interactions, and cultural meanings associated with the experimental activities [@gowlland2019]. Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction. One potential application of ethnographic methods in experimental archaeology of stone artifacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.

@@

The comparative study and large-scale synthesis of variation data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology. Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the highest file size compared with the other two. As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library that was originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.
